Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-18 article.tex
add averaging discussion in experiments + corr ack

As the first plot in Figure~\ref{fig:filtered-signal} illustrates, due to
light scattering artifacts that ordinarily affects the quality of the
recording~\citep{lichtman2011big}, the raw fluoresence signal is very noisy.
Accordingly, the first step of our pipeline is to smoothen the signal, using a low-pass filter for
filtering out high frequency noise, applying one of the following designs and
resulting in the second plot in the figure:
\begin{align}
% symmetrical median filter
x^t_i &\leftarrow x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
x^t_i &\leftarrow 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i.
\label{eq:weighted-asymetric-median}
\end{align}

Considering the short delay of communication between neurons and the slow
decay of fluorescence, the second step of our pipeline consists in identifying
neuron spikes -- characterized by high frequencies -- by filtering out low
frequencies through a high-pass filter, resulting in third plot of the figure:
\begin{align} % an asymmetrical discrete derivative filter 
x^{t,d}_{i} &\leftarrow x^{t}_i - x_{t-1}^i. \label{eq:high-pass-filter} 
\end{align}

Once the low- and and high-pass filters have been applied, the remaining signal
corresponds to the discrete derivative of the fluorescence time series. To
further filter out small signal variations, mostly due to noise, the third step
consists in a  hard-threshold filter, yielding the fourth plot in the figure:
\begin{align}
x^{t,d}_i &\leftarrow x^{t,d}_i \mathbb{1}(x^{t,d}_i \geq \tau) \text{ with } \tau > 0.
\end{align}

At this step, the processed signal only contains clean spikes. However, when a
large part of the network is firing, detecting pairwise neuron interactions is a
difficult task. \textcolor{red}{G: say that neurons may fire because of global
high activity, without any actual cause -- see generator.}
Accordingly, in order to reduce false interaction detection,
peak intensities at time $t$ are regularized by the global activity of all
neurons, thereby reducing the importance of peaks identified during global high activity periods:
\begin{align}
 x^{t,w}_i  &\leftarrow (x^{t,d}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t,d}_j}}.
\end{align}

\textcolor{red}{G: Change notations for $f(x^t) = ...$, $g(x^t) = ...$}

@@

At first sight, it appears on Tab.\ref{tab:tab1} that partial
correlation outperforms other methods. Two main reasons may explain this.
First, the quality of the infered network seems related to the conditioning
part of the method. In every method except with partial correlation, indirect
links are confounded with direct links.  Secondly, the gaussian assumption
seems verified and thus linear method appears to be competitive, e.g. the
method correlation already gives good results, with a non-linear method such
as GENIE3.\\

@@

Observing Tab.\ref{tab:tab1}, it appears that
best methods only infer the \textbf{undirected} network by giving the same
weights to both edges $(i,j)$ and $(j,i)$. This is quite surprising
considering the goal of the challenge, namely predicted directed connections
between neurons. Besides, if the true network undirected, i.e. for all links
$(i,j)$, back links are added, is the infered network then the score obtained
according to the ROC AUC metric is very high ($\sim 0.9956$ on
\textit{normal-1}). Therefore, an undirected infered network seems sufficient
to maximize the ROC AUC score without trying to identify the link direction.
Indeed, predicting perfectly directivity only yields a small gain ($\sim
0.005$ on \textit{normal-1}) which is is not very rewarding. In contrast, the
same experiment with the precision recall metric rewards less the undirected
network and the gain margin in predicting the link directivity is large enough
to be interesting. For these reason, it seems that the metric used in the
contest, i.e. the ROC AUC, does not seem consistent with the goal of
predicting directed connections between neurons. Another metric, such as the
AUC PRC, is maybe more appropriate to assess the quality of the
\textbf{directed} infered network.

@@

A. Joly and G. Louppe are research fellows of
the FNRS, Belgium.  A. Sutera is  is recipient of
a F.R.I.A. fellowship of F.R.S.-FNRS, Belgium.
This work is supported by PASCAL2 and the IUAP DYSCO, initiated by the
Belgian State, Science Policy Office.

>>>>

As the first plot in Figure~\ref{fig:filtered-signal} illustrates, due to
light scattering artifacts that ordinarily affects the quality of the
recording~\citep{lichtman2011big}, the raw fluoresence signal is very noisy.
Accordingly, the first step of our pipeline is to smoothen the signal, using a low-pass filter for
filtering out high frequency noise, applying one of the following designs and
resulting in the second plot in the figure:
\begin{align}
% symmetrical median filter
f_1(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i.
\label{eq:weighted-asymetric-median}
\end{align}

Considering the short delay of communication between neurons and the slow
decay of fluorescence, the second step of our pipeline consists in identifying
neuron spikes -- characterized by high frequencies -- by filtering out low
frequencies through a high-pass filter, resulting in third plot of the figure:
\begin{align} % an asymmetrical discrete derivative filter 
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i. \label{eq:high-pass-filter} 
\end{align}

Once the low- and and high-pass filters have been applied, the remaining signal
corresponds to the discrete derivative of the fluorescence time series. To
further filter out small signal variations, mostly due to noise, the third step
consists in a  hard-threshold filter, yielding the fourth plot in the figure:
\begin{align}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0.
\end{align}

At this step, the processed signal only contains clean spikes. However, when a
large part of the network is firing, detecting pairwise neuron interactions is a
difficult task. \textcolor{red}{G: say that neurons may fire because of global
high activity, without any actual cause -- see generator.}
Accordingly, in order to reduce false interaction detection,
peak intensities at time $t$ are regularized by the global activity of all
neurons, thereby reducing the importance of peaks identified during global high activity periods:
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}.
\end{align}

The full signal processing of our approach is defined by
\begin{align}
w(h(g(f_k(x_i^t)))) 
\end{align}
where $f_k$ is one of the low-pass filter designs.

@@

At first sight, it appears on Tab.\ref{tab:tab1} that partial
correlation outperforms other methods. Two main reasons may explain this.
First, the quality of the infered network seems related to the conditioning
part of the method. In every method except with partial correlation, indirect
links are confounded with direct links.  Secondly, the gaussian assumption
seems verified and thus linear method appears to be competitive, e.g. the
method correlation already gives good results, with a non-linear method such
as GENIE3.

% Averaging

It also appears that the averaging over the low-pass filter designs and the
threshold values increases significantly the quality of the infered network
according to both metrics.

@@

Observing Tab.\ref{tab:tab1}, it appears that best methods only infer the
\textbf{undirected} network by giving the same weights to both edges $(i,j)$
and $(j,i)$. This is quite surprising considering the goal of the challenge,
namely predicted directed connections between neurons. Besides, if the true
network undirected, i.e. for all links $(i,j)$, back links are added, is the
infered network then the score obtained according to the ROC AUC metric is
very high ($\approx 0.9956$ on \textit{normal-1}). Therefore, an undirected
infered network seems sufficient to maximize the ROC AUC score without trying
to identify the link direction. Indeed, predicting perfectly directivity only
yields a small gain ($\approx 0.005$ on \textit{normal-1}) which is is not
very rewarding. In contrast, the same experiment with the precision recall
metric rewards less the undirected network ($\approx 0.789$ on
\textit{normal-1}) and the gain margin in predicting the link directivity is
large enough to be interesting ($\approx 0.211$ on \textit{normal-1}). For
these reason, it seems that the metric used in the contest, i.e. the ROC AUC,
does not seem consistent with the goal of predicting directed connections
between neurons. Another metric, such as the AUC PRC, is maybe more
appropriate to assess the quality of the \textbf{directed} infered network.

@@

A. Joly and G. Louppe are research fellows of
the FNRS, Belgium.  A. Sutera is recipient of
a F.R.I.A. fellowship of F.R.S.-FNRS, Belgium.
This work is supported by PASCAL2 and the IUAP DYSCO, initiated by the
Belgian State, Science Policy Office.
